---
title: "`wholebrain` - a basic (for now) computer vision library for R"
author: "Simon Garnier - <a href='https://twitter.com/sjmgarnier'>@sjmgarnier</a>"
date: "`r Sys.Date()`"
output:
  html_document:
    css: custom.css
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    number_sections: yes
    theme: cosmo
    highlight: pygments
    toc: yes
  pdf_document:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Vignette Title} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

---

# Introduction

[R](http://www.r-project.org) is by far my favorite tool to process the data 
from my lab's experiments, generate beautiful and informative graphs from them, 
and compute all the statistics that I need for scientific reports and articles.
However it lacks one important toolset for my work: the ability to process 
images and videos. I can always rely on a [Matlab](http://www.mathworks.com/products/matlab/)
or a [Python](https://www.python.org/) script for this, but I would prefer 
handling my analysis workflow entirely with R rather than depending on mutiple
tools and languages for this.

`videoplayR` is my attempt at solving this problem by creating a computer vision
toolbox for R, based on the [OpenCV](http://opencv.org/) C++ library. In its 
current state, `videoplayR` is very limited when compared to the libraries 
provided by Matlab and Python, but (1) it is functional enough to import images 
from image files and videos and perform some basic processing on them, and (2) 
it is a very young package that I hope will start growing quickly. 

This vignette is intended to show how to install and run `videoplayR`, and to 
demonstrate its current computer vision capabilities. If you run into a bug or 
would like to suggest improvements, please head to the package's Github 
[repository](https://github.com/swarm-lab/videoplayR) and submit a new issue at
[https://github.com/swarm-lab/videoplayR/issues](https://github.com/swarm-lab/videoplayR/issues).
I also welcome the help of R developers to improve `videoplayR`. The best way to 
do this is to fork the repository and submit pull requests for your changes.

[Back to top]("#")

---

# Installation

`videoplayR` is not available on CRAN, at least for now. The reason is that it 
has external dependencies (OpenCV) that are difficult to include in the package
directly. Until someone finds a convenient, multi-platform solution to ship 
OpenCV source files with the package, it will remain available through Github 
only.

Unfortunately `videoplayR` is also not (yet) available for Windows. I have zero 
experience preparing Rcpp-based packages for Windows and currently I do not have 
the time and resources to learn how to compile packages for Windows. If you are 
a R/Rcpp developer with experience preparing packages for Windows, and you want 
to help bring `videoplayR` to Windows, you can head to `videoplayR` [Github repository](https://github.com/swarm-lab/videoplayR), clone it and send a pull 
request once you have figured out how to install OpenCV and compile the package 
on Windows. 

The good news is that it is not very difficult to install the required 
dependencies on Mac and Linux. A few command lines in the terminal are enough to
prepare your computer for and compile `videoplayR`. The steps to install the 
package on your computer are detailed in the sections below. 

## Installing `opencv`

### Mac

The easiest way to install all the external dependencies for `videoplayR` is to 
use a package manager such as [Homebrew](http://brew.sh/) or [MacPorts](https://www.macports.org).
I personally favor Homebrew, but MacPorts will also install the dependencies 
just fine. 

**- Homebrew -**

To install Homebrew, just head to your favorite terminal app (OSX Terminal for
instance) and run the following command line: 

```{bash, eval=FALSE} 
ruby -e "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/master/install)"
```

Then, install OpenCV 3.0 by simply typing the following command lines in the 
terminal:

```{bash, eval=FALSE}
brew tap homebrew/science
brew install pkg-config opencv3
brew link opencv3 --force
```

**- MacPorts -**

You can install MacPorts by following the instructions provided on its website
at [https://www.macports.org/install.php](https://www.macports.org/install.php).

Once MacPorts is installed, you will be able to install OpenCV 3.0 by simply 
typing the following command line in the terminal (you will need the 
administrator password):

```{bash, eval=FALSE}
sudo port install opencv
```

### Linux

On Linux, you should be able to install the development files of OpenCV using 
your usual package manager. 

**- apt based systems (Ubuntu, Debian, etc.) -**

```{bash, eval = FALSE}
sudo apt-get install libopencv-dev
```

**- yum based systems (Red Hat, Fedora, etc.) -**

```{bash, eval = FALSE}
sudo yum install opencv-devel
```

**- Arch Linux -**

```{bash, eval = FALSE}
sudo pacman -S opencv
```

## Installing `videoplayR` from Github

Once OpenCV and its dependencies are installed, you can open your favorite R 
development environment and enter the following commands in the R console in 
order to install the development version of `videoplayR` from its Github 
repository. 

```{r, eval=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
}

devtools::install_github("swarm-lab/videoplayR")

library(videoplayR)
```

That's it! `videoplayR` should now be ready to work on your system. If you 
experience any trouble during the installation process, please report it on the 
issue page of the package's Github repository at 
[https://github.com/swarm-lab/videoplayR/issues](https://github.com/swarm-lab/videoplayR/issues).

[Back to top]("#")

---

# Loading images

The core of the `videoplayR` package is the `vpImage` object class. A `vpImage` 
object is a wrapper around an OpenCV image matrix that can be manipulated 
directly from within R. There are three ways to create `vpImage` objects: from 
image files, from video frames, and from R matrices and arrays. I will 
demonstrate these different methods in the rest of this section.

**Note that `vpImage` objects are not persistent objects like any normal R 
object. They cannot be saved for reuse in a different session.**

```{r, echo=FALSE, message=FALSE}
library(videoplayR)
```

## From image files

The `readImg` function will read an image file and return a `vpImage` object. In
the example below, I use an image file included in the package. If you prefer to
use an image from your computer, just replace the filename with a character 
string representing the path to your image file.

```{r}
filename <- system.file("sample_img/SampleVideo_1080x720_5mb.png", package = "videoplayR")
img1 <- readImg(filename)
```

A `vpImage` object has a couple of attributes that represent the type of image 
stored inside (binary, grayscale, RGB, or numeric) and the dimensions of the 
picture (width and height in pixels, and number of channels if applicable). You 
can access these attributes as follows. 

```{r}
img1$type
img1$dim
```

Finally, you can display the `vpImage` object using the `imshow` function.

```{r, fig.height=4.5}
imshow(img1)
```

Since `imshow` display the image in a regular R graphics device, you can plot 
lines, dots, etc., over the image display. Note that the bottom left of the 
image corresponds to the {1, 1} coordinates.

```{r, fig.height=4.5}
imshow(img1)
points(c(500, 530), c(505, 505), col = "red", pch = 19, cex = 2)
lines(c(1, img1$dim[2]), c(1, img1$dim[1]), col = "red", lwd = 8)
lines(c(1, img1$dim[2]), c(img1$dim[1], 1), col = "red", lwd = 8)
```

## From videos frames

As its name suggests, `videoplayR` can also read videos. It cannot (yet) play
them, but it can read video frames and import them as `vpImage` objects. Videos
are accessed through a special object class: `vpVideo`. A `vpVideo` object is a 
wrapper around an OpenCV video reader object and it can be created using the 
`readVid` function. 

```{r}
filename <- system.file("sample_vid/SampleVideo_1080x720_5mb.mp4", package = "videoplayR")
vid1 <- readVid(filename)
```

A `vpVideo` object has three attributes that represent the length of the video 
(in frames), the framerate of the video (in frames per second), and the dimesions 
of the video (width and heigth in pixels). You can access these attributes as follows. 

```{r}
vid1$length
vid1$fps
vid1$dim
```

You can grab any frame in the video as a `vpImage` object by using the `getFrame` 
function as follows. 

```{r, fig.height=4.5}
img2 <- getFrame(vid1, 100)
imshow(img2)
```

## From your webcam

Since version 0.3.1, `videoplayR` can also access the video stream from cameras
connected to your computer, and import frames as `vpImage` objects. Camera 
streams are accessed through a special object class: `vpStream`. A `vpStream` 
object is a wrapper around an OpenCV video reader object and it can be created 
using the `readStream` function. In the following example, we will open a stream
from your default webcam.

```{r}
stream1 <- readStream(0)  # 0 is the index of your default webcam. Use higher 
                          # numbers to access other cameras connected to your 
                          # computer. 
```

A `vpStream` object has one attribute that represents the dimesions of the video 
stream (width and heigth in pixels). You can access this attribute as follows. 

```{r}
stream1$dim
```

You can grab the current frame in the stream as a `vpImage` object by using the 
`nextFrame` function as follows. 

```{r, echo=FALSE}
Sys.sleep(1)
```

```{r, fig.height=4.5, cache=TRUE}
img3 <- nextFrame(stream1)
imshow(img3)
```

Finally, when you are done using your camera, you can release (or close) the 
stream using the `release` function as follows. 

```{r}
release(stream1)
```

## From R matrices and arrays

Fundamentally, a digital image is nothing more than a matrix of numbers. Therefore
I added a couple of functions that convert R matrices and arrays (3D arrays only)
into `vpImage` objects, and vice versa. 

The `r2img` function takes an R matrix or 3D array and converts it to a `vpImage`
object. For instance, in the following code chunk, I create a gradient matrix 
that I convert to a `vpImage` object and then display using the `imshow` function. 

```{r}
gradient <- function(x1, x2) { x1 + x2 }
x <- 0:127
mat1 <- gradient(outer(rep(1, length(x)), x), x)
img4 <- r2img(mat1)
imshow(img4)
```

On the other hand, the `img2r` function takes a `vpImage` object and converts it 
to an R matrix (for binary, grayscale and 1-channel numeric objects) or a 3D 
arrays (for RGB and 3-channels numeric objects). You can then manipulate the 
matrix or array like you would do with any other regular R matrix or array, and 
then convert it back to a `vpImage` object, like in the example below. 

```{r, fig.height=4.5}
mat2 <- img2r(img1)
mat2[80:640, 250:650, ] <- 255
img5 <- r2img(mat2)
imshow(img5)
```

Note that a number of operations can be applied to a `vpImage` object directly 
without requiring a conversion to an R object first. I will detail these 
operations in the next section. 

[Back to top]("#")

---

# Operations on images

Currently `videoplayR` provides only a small fraction of the image processing 
capabilities of OpenCV. The number of functions will increase little by little
as my needs for more advanced features develop or as other R developers provide
their own code to be included in the package (hint: pull requests to the project
repository are very, very welcome: 
[https://github.com/swarm-lab/videoplayR/pulls](https://github.com/swarm-lab/videoplayR/pulls))

Nevertheless, `videoplayR` has enough capabilities to get anybody started with 
basic image processing and, with a bit of creativity, to achieve more advanced 
results such as simple object detection and tracking. I review the image processing
functions provided by `videoplayR` in the rest of this section.

## Thresholding

The most common way to generate binary images (images with pixel values that are 
either zeros or ones) is to use a method called thresholding: all the pixels with 
a value above a given threshold are set to the same value (one for instance), and 
all the pixels with a value below the threshold are set to another one (zero for 
instance).

The `thresholding` function in the `videoplayR` package does just that. It takes
three arguments: the original image to binarize (`image =`), a threshold value
(`thres = `), and the type of thresholding (`type =`). If `type = "binary"` then 
the pixels above `thres` are set to one, and the pixels below are set to zero.

```{r, fig.height=4.5}
bin1 <- thresholding(image = img1, thres = 128, type = "binary")
imshow(bin1)
```

If `type = "inverted"` then the pixels above `thres` are set to zero, and the 
pixels below are set to one

```{r, fig.height=4.5}
bin2 <- thresholding(img1, 128, type = "inverted")
imshow(bin2)
```

Note that if the original image has multiple channels (RGB image or 3-channels 
numeric image), it is first "flattened" to one channel automatically by the 
`thresholding` function. Image flattening is explained in the following section.

## Flattening 

Flattening an image is the process of reducing its depth, that is the number of 
channels that it is made of. A typical example of flattening is when one wants to
transform an RGB image (3 channels: Red, Green, Blue) to a grayscale image (1 channel).

`videoplayR` offers a simple way to flatten a 3-channels image by using the 
`ddd2d` function (depth 3 to depth 1). Hereafter it is used to convert an RGB 
image to grayscale. 

```{r, fig.height=4.5}
gray1 <- ddd2d(img1)
imshow(gray1)
```

You can also expand a 1-channel image to 3 channels using the `d2ddd` function 
(depth 1 to depth 3). 

```{r, fig.height=4.5}
gray2 <- d2ddd(gray1)
imshow(gray2)
```

This will not restore the colors of the original RGB image (this information is 
lost), but it can be a practical way to make 3-channels mask or intensity images
as I will demonstrate in the next section about blending images.

## Blending

Blending two images is similar to doing element-wise operations on two R matrices 
or arrays. Each pixel of an image A is combined with the corresponding pixel of 
another image B through a mathematical operation: they can be added together, 
subtracted from each other, multiplied with each other, or divided by each other.

When using `vpImage` objects, it is possible to blend two images together without
having to convert them to R matrices or arrays first. This is done using the 
`blend` function that works taking two `vpImage` objects (`image1 = ` and `image2 =`)
and combining them using one of four possible operations (`+`, `-`, `*`, `/`). 

In the following example, I create an uniform 3-channels image (`lum`) with all 
pixels set to 50. I can then add or subtract this uniform image from the first 
`vpImage` object that we had created earlier in order to adjust the overall 
luminosity of the picture. 

```{r, fig.height=9}
mat <- matrix(50, nrow = img1$dim[1], ncol = img1$dim[2])
lum <- d2ddd(r2img(mat))

par(mfrow = c(3, 1))

imshow(blend(img1, lum, "-"))
text(x = 1240, y = 40, labels = "Intensity - 50", adj = 1, cex = 3, col = "white")

imshow(img1)
text(x = 1240, y = 40, labels = "Original", adj = 1, cex = 3, col = "black")

imshow(blend(img1, lum, "+"))
text(x = 1240, y = 40, labels = "Intensity + 50", adj = 1, cex = 3, col = "black")
```

We can also use the `blend` function to mask part of the image, that is turn the 
pixels to black (i.e. to zero) in the masked part. Hereafter I create a 3-channels
mask image (`mask`) and combine it with our example image using the multiplication
operation. The pixels set to zero in the mask image will turn the corresponding 
pixels in the example image to black after blending, while pixels set to one in 
the mask image will not modify the corresponding pixels in the example image. 

```{r, fig.height=4.5}
mat <- matrix(0, nrow = img1$dim[1], ncol = img1$dim[2])
mat[80:640, 250:650] <- 1
mask <- d2ddd(r2img(mat))

imshow(blend(img1, mask, "*"))
```

By combining blending operations with creative masking and luminance images, you 
can achieve pretty complex image transformations.

## Simple background extraction

In my work, I often need to separate objects in the foreground from the image 
background. It is fairly easy to do if you have a good image of the background 
in the absence of the objects. However if you do not possess a clean picture of 
the background (for instance when there are constantly moving objects in the 
frame), you need to reconstruct the background image. A simple way to do so is 
to estimate for each pixel its most frequent value throughout a video, assuming
that each pixel is not visited too often by the objects in the video. 

The `backgrounder` function in `videoplayR` is my attempt at providing a tool to
automatically reconstruct the background image of a video. This functions takes 
four arguments: 

+ a `vpVideo` object (`video =`).
+ the number of frames to sample for reconstructing the background (`n =`). Larger 
numbers give more accurate results but cause the function to run slower. 
+ the method to extract the background (`method =`). If `method = "mean"`, the 
background image will correspond to the average value of the sampled frames. It 
is the fastest method, but not the most accurate. If `method = "median"`, the 
background image will correspond to the median value of the sampled frames. It 
is more accurate, but also slower. 
+ whether the background should be returned as a color image (`color = TRUE`, 
slower) or as a grayscale image (`color = FALSE`, faster).

In the following example, I reconstruct the background of a video in which people
are normally passing through. The first figure below shows a frame containing a 
person. The second figure shows the reconstructed background image.

```{r}
filename <- system.file("sample_vid/Walk3.mp4", package = "videoplayR")
vid2 <- readVid(filename)
imshow(getFrame(vid2, 125))
```

```{r}
bg <- backgrounder(vid2, n = 100, method = "median", color = TRUE)
imshow(bg)
```

## Simple blob detection

Another important task that I often need to perform is detect the location of 
objects in an image, for instance after removing its background. This task is 
often called blob detection.

`videoplayR` implements a very simple blob detection function called `blobDetector`. 
It will find white blobs on a black background and will return their locations 
in the image as well as some information about their shapes (orientation of the 
major axis, lengths of the major and minor axis in pixels, surface area in pixels). 

In the example below, I demonstrate how to detect objects on a uniform background 
using the `blobDetector` function. 

First, I load and display the test image.

```{r}
filename <- system.file("sample_img/dots.jpg", package = "videoplayR")
img6 <- readImg(filename)
imshow(img6)
```

The objects in this image are darker than the background. The first step for letting 
the computer detect them is to convert the color image to a binary one, with the 
objects to detect appearing white (value of one) and the background black (value 
of zero). 

```{r}
bin3 <- thresholding(img6, 200, "inverted")
imshow(bin3)
```

We can now pass this binary image to the `blobDetector` function that will return 
a data frame in which each row corresponds to a specific object in the image. 

```{r}
blobs <- blobDetector(bin3)
head(blobs)
```

Each column of the data frame corresponds to a property of each blob:

+ "id": identity of the blob.
+ "x": x coordinate of the center of the ellipse approximating the shape of the blob.
+ "y": y coordinate of the center of the ellipse approximating the shape of the blob.
+ "alpha": orientation in degrees of the major axis of the ellipse.
+ "major": length of the major axis of the ellipse.
+ "minor": length of the minor axis of the ellipse.
+ "area": area of the polygon containing the blob.

Finally, we can plot the result of the blob detection algorithm over the original
image. This requires the installation of the `plotrix` package to add ellipses 
showing the minor and major axes of each object.

```{r, message=FALSE}
require(plotrix)
imshow(img6)
points(y ~ x, data = blobs, col = "red", pch = 20)
draw.ellipse(blobs$x, blobs$y, blobs$major / 2, blobs$minor / 2, blobs$alpha, border = "red")
```

The estimated ellipses are not perfect for every object, but the center of mass 
is pretty accurate in each case. 

[Back to top]("#")

---














